---
knit: bookdown::preview_chapter
---

# Introduction {#intro}

This section should contain a summary of, and the outline of any theory to, the method that you’re are describing. It should also outline the major procedures involved in the protocol.

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab}
knitr::kable(head(iris, 20), caption = 'Here is a nice table!', booktabs = TRUE)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].



Segmentation is critical to image quantification.  In the context of brain mapping, the question of where (localization) is as important as quantity.  

image classification 

data science bowl cardiac 

importance of challenges 

imagine the future: cloud-based databases of labeled data that let us ID patterns in medical images

public resources - what tools are needed for this scenario?

* tissue based on intensity 

* learning from training 

* identify networks


---- like online quality assessment etc to allow rescanning etc

---- longitudinal stability wrt hardware / software upgrades

---- how validate gradient-induced distortion correction?

---- performance relative to other methods - esp on older subjects


